---
title: "Testing the integration of vega-lite in RMarkdown documents"
author: "Miles Benton"
date: "30/05/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up the environment

```{r, message=FALSE, warning=FALSE}
# require(altair)       # this is the r wrapper for the python package
require(vegawidget)
require(reticulate)   # allows us to run python alongside R in dedicated chunks
reticulate::use_condaenv("r-reticulate")    # dedicated conda env with appropriate packages installed
matplotlib <- import("matplotlib")          # these two lines allow the plotting of graphs using
matplotlib$use("Agg", force = TRUE)         # the matplotlib library for python
```

# testing `matplotlib` plotting in RMarkdown

```{python}
import pandas as pd
import matplotlib.pyplot as plt
df=pd.DataFrame([[1, 2], [3, 4], [4, 3], [2, 3]])
fig = plt.figure(figsize=(14,8))
for i in df.columns:
    ax=plt.subplot(2,1,i+1) 
    df[[i]].plot(ax=ax)
    print(i)

plt.show()
```

# create vega-lite plot using `altair` and display in document

```{python}
import altair as alt
from vega_datasets import data

source = data.population.url

chart = alt.Chart(source).mark_boxplot(extent='min-max').encode(
  x='age:O',
  y='people:Q'
)

chart.save('figures/chart.html')
```

```{r}
print(py$chart)
```

### The above chart should be rendered below using `iframe` tags

Above we saved the plot out to an html file (`chart.html`). Using `iframe` tags we can then embed this back into the RMarkdown document.

Here is the code: `<iframe width="600" height="400" src="figures/chart.html" frameborder="0"></iframe>`

Below is the result:

<iframe width="600" height="400" src="figures/chart.html" frameborder="0"></iframe>

# Testing a second `iframe` embed

```{python}
import pandas as pd
data = pd.DataFrame({'a': list('CCCDDDEEE'),
                     'b': [2, 7, 4, 1, 2, 6, 8, 4, 7]})
chart2 = alt.Chart(data).mark_bar(color='firebrick').encode(
    alt.Y('a', title='category'),
    alt.X('average(b)', title='avg(b) by category')
)
chart2.save('figures/chart2.html')
```

### The above chart should be rendered below using `iframe` tags:

<iframe width="600" height="135" src="figures/chart2.html" frameborder="0"></iframe>

# using `vegawidget` in R

This time we'll use R to create a vega-lite plot, specifically the `vegawidget` package:

```{r}
spec_mtcars <-
  list(
    `$schema` = vega_schema(), # specifies Vega-Lite
    description = "An mtcars example.",
    data = list(values = mtcars),
    mark = "point",
    encoding = list(
      x = list(field = "wt", type = "quantitative"),
      y = list(field = "mpg", type = "quantitative"),
      color = list(field = "cyl", type = "nominal")
    )
  ) %>% 
  as_vegaspec()

spec_mtcars
```

This package has several advantages in terms of ability to integrate with Shiny, will look into this further.

With a little bit of tweaking it looks like boxplots are supported in `vegawidget`, this is an advantage over the R package `vegalite` and the R implementation of `altair` which both seem to be lagging behind on an older version of vega-lite.

Example of a boxplot generated using R:

```{r}
population <- jsonlite::fromJSON("data/population.json")

  list(
    `$schema` = vega_schema(), # specifies Vega-Lite
    description = "An boxplot example.",
    data = list(values = population),
    mark = list(type = "boxplot", extent = 'min-max'),
    encoding = list(
      x = list(field = "age", type = "nominal"),
      y = list(field = "people", type = "quantitative")
    )
  ) %>% 
  as_vegaspec()
```

## Faceted chart

```{r, warning=FALSE}
facet_chart <-
  list(
    `$schema` = vega_schema(),
    data = list(values = data_seattle_daily),
    facet = list(
      row = list(
        field = "date", 
        type = "ordinal", 
        timeUnit = "utcyear",
        title = "Faceted chart"
      )
    ),
    spec = list(
      height = 100,
      mark = "bar",
      encoding = list(
        x = list(
          field = "date", 
          type = "ordinal", 
          timeUnit = "utcmonth",
          title = "month(date)"
        ),
        y = list(
          field = "temp_max",
          type = "quantitative",
          aggregate = "max",
          title = "max(temp_max)"
        )
      )
    )
  ) %>%
  as_vegaspec()

facet_chart
```

```{python}
import altair as alt
from vega_datasets import data

source = data.iris()

base = alt.Chart(source)

xscale = alt.Scale(domain=(4.0, 8.0))
yscale = alt.Scale(domain=(1.9, 4.55))

area_args = {'opacity': .3, 'interpolate': 'step'}

points = base.mark_circle().encode(
    alt.X('sepalLength', scale=xscale),
    alt.Y('sepalWidth', scale=yscale),
    color='species',
)

top_hist = base.mark_area(**area_args).encode(
    alt.X('sepalLength:Q',
          # when using bins, the axis scale is set through
          # the bin extent, so we do not specify the scale here
          # (which would be ignored anyway)
          bin=alt.Bin(maxbins=20, extent=xscale.domain),
          stack=None,
          title=''
         ),
    alt.Y('count()', stack=None, title=''),
    alt.Color('species:N'),
).properties(height=60)

right_hist = base.mark_area(**area_args).encode(
    alt.Y('sepalWidth:Q',
          bin=alt.Bin(maxbins=20, extent=yscale.domain),
          stack=None,
          title='',
         ),
    alt.X('count()', stack=None, title=''),
    alt.Color('species:N'),
).properties(width=60)

scatterhist = top_hist & (points | right_hist)
scatterhist.save("figures/scatterhist.html")
```

<iframe width="700" height="480" src="figures/scatterhist.html" frameborder="0"></iframe>

## maps

```{python}
import altair as alt
from vega_datasets import data

counties = alt.topo_feature(data.us_10m.url, 'counties')
source = data.unemployment.url

usmap = alt.Chart(counties).mark_geoshape().encode(
    color='rate:Q'
).transform_lookup(
    lookup='id',
    from_=alt.LookupData(source, 'id', ['rate'])
).project(
    type='albersUsa'
).properties(
    width=500,
    height=300
)

usmap.save('figures/usmap.html')
```

<iframe width="700" height="350" src="figures/usmap.html" frameborder="0"></iframe>

```{python}
import altair as alt
from vega_datasets import data

source = alt.topo_feature(data.world_110m.url, 'countries')

base = alt.Chart(source).mark_geoshape(
    fill='#666666',
    stroke='white'
).properties(
    width=300,
    height=180
)

projections = ['equirectangular', 'mercator', 'orthographic', 'gnomonic']
charts = [base.project(proj).properties(title=proj)
          for proj in projections]

worldproj = alt.concat(*charts, columns=2)
worldproj.save("figures/worldproj.html")
```

<iframe width="700" height="490" src="figures/worldproj.html" frameborder="0"></iframe>

```{python}
import altair as alt
from vega_datasets import data

# Since the data is more than 5,000 rows we'll import it from a URL
source = data.zipcodes.url

zipcode = alt.Chart(source).transform_calculate(
    "leading digit", alt.expr.substring(alt.datum.zip_code, 0, 1)
).mark_circle(size=3).encode(
    longitude='longitude:Q',
    latitude='latitude:Q',
    color='leading digit:N',
    tooltip='zip_code:N'
).project(
    type='albersUsa'
).properties(
    width=650,
    height=400
)

zipcode.save("figures/zipcode.html")
```

<iframe width="800" height="475" src="figures/zipcode.html" frameborder="0"></iframe>

## basic interaction

```{python}
import altair as alt
from vega_datasets import data

source = data.cars()

intplot = alt.Chart(source).mark_circle().encode(
    x='Horsepower',
    y='Miles_per_Gallon',
    color='Origin',
).interactive()

intplot.save("figures/intplot.html")
```

<iframe width="600" height="400" src="figures/intplot.html" frameborder="0"></iframe>

# something a bit more advanced...

## points and bars (Python)

```{python}
import altair as alt
from vega_datasets import data

source = data.cars()

brush = alt.selection(type='interval')

points = alt.Chart(source).mark_point().encode(
    x='Horsepower:Q',
    y='Miles_per_Gallon:Q',
    color=alt.condition(brush, 'Origin:N', alt.value('lightgray'))
).add_selection(
    brush
)

points.save('figures/points.html')

bars = alt.Chart(source).mark_bar().encode(
    y='Origin:N',
    color='Origin:N',
    x='count(Origin):Q'
).transform_filter(
    brush
)

bars.save('figures/bars.html')

# points & bars

barpointchart = alt.vconcat(points, bars)
barpointchart.save('figures/barpointchart.html')
```

<iframe width="700" height="495" src="figures/barpointchart.html" frameborder="0"></iframe>

## weather

```{python}
import altair as alt
from vega_datasets import data

source = data.seattle_weather()

scale = alt.Scale(domain=['sun', 'fog', 'drizzle', 'rain', 'snow'],
                  range=['#e7ba52', '#a7a7a7', '#aec7e8', '#1f77b4', '#9467bd'])
color = alt.Color('weather:N', scale=scale)

# We create two selections:
# - a brush that is active on the top panel
# - a multi-click that is active on the bottom panel
brush = alt.selection_interval(encodings=['x'])
click = alt.selection_multi(encodings=['color'])

# Top panel is scatter plot of temperature vs time
points = alt.Chart().mark_point().encode(
    alt.X('monthdate(date):T', title='Date'),
    alt.Y('temp_max:Q',
        title='Maximum Daily Temperature (C)',
        scale=alt.Scale(domain=[-5, 40])
    ),
    color=alt.condition(brush, color, alt.value('lightgray')),
    size=alt.Size('precipitation:Q', scale=alt.Scale(range=[5, 200]))
).properties(
    width=550,
    height=300
).add_selection(
    brush
).transform_filter(
    click
)

# Bottom panel is a bar chart of weather type
bars = alt.Chart().mark_bar().encode(
    x='count()',
    y='weather:N',
    color=alt.condition(click, color, alt.value('lightgray')),
).transform_filter(
    brush
).properties(
    width=550,
).add_selection(
    click
)

weather = alt.vconcat(
    points,
    bars,
    data=source,
    title="Seattle Weather: 2012-2015"
)
weather.save("figures/weather.html")
```

<iframe width="770" height="550" src="figures/weather.html" frameborder="0"></iframe>

## points and bars (`vegawidget`)

!! Below is not working at this stage due to a conflict in versions between Python and R versions of altair and their versions of vegalite.

```{r, eval=FALSE}
scale <- alt$Scale(
  domain = list("sun", "fog", "drizzle", "rain", "snow"),
  range = list("#e7ba52", "#a7a7a7", "#aec7e8", "#1f77b4", "#9467bd")
)

color <- alt$Color("weather:N", scale = scale)

# We create two selections:
# - a brush that is active on the top panel
# - a multi-click that is active on the bottom panel
brush <- alt$selection_interval(encodings = list("x"))
click <- alt$selection_multi(encodings = list("color"))

# Top panel is scatter plot of temperature vs time
points <- 
  alt$Chart()$
  mark_point()$
  encode(
    x = alt$X(
      "date:T", 
      timeUnit = "monthdate", 
      axis=alt$Axis(title="Date")
    ),
    alt$Y(
      "temp_max:Q",
      axis = alt$Axis(title = "Maximum Daily Temperature (C)"),
      scale = alt$Scale(domain = list(-5, 40))
    ),
    color = alt$condition(brush, color, alt$value("lightgray")),
    size = alt$Size("precipitation:Q", scale = alt$Scale(range = list(5, 200)))
  )$
  properties(width = 600, height = 300, selection = brush)$
  transform_filter(click$ref())

# Bottom panel is a bar chart of weather type
bars <- 
  alt$Chart()$
  mark_bar()$
  encode(
    x = "count(weather)",
    y = "weather:N",
    color = alt$condition(click, color, alt$value("lightgray"))
  )$
  transform_filter(brush$ref())$
  properties(width = 600, selection = click)

chart <- 
  (points & bars)$ 
  properties(
    data = vega_data$seattle_weather(),
    title = "Seattle Weather: 2012-2015"
  )  

chart
```

